Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-11 paper/paper.tex
Added visualizations of laziness and strictness.

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. Let $wn :
\widehat{State} \times Store \to State$ be the metafunction that adds
a store to a storeless state. Formally,

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\} \\
             \msto' &= \bigcup\{\msto^c \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\}
\end{align*}

@@

\begin{align*}
\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\msto,\mkont} &\machstep
\ev{\mexpi{0},\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
\\
\co{\kar{\mexp,\mlab},\mval,\msto} & \machstep
\ev{\mexp,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
\\
\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
\\
\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
\\
\ap[_\mlab]{\clos{\mvar,\mexp},\mlab^a, \msto,\mkont} & \machstep
\ev{\mexp,\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
\\
\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
\end{align*}

@@

The first problem that we realized with the AAM approach was that much
of the non-determinism is spurious. For example, in a function
application,
\begin{alltt}
(f x y z w)
\end{alltt}
where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt x}, {\tt y}, {\tt z} and
{\tt w}, only to be joined back together in {\tt f}'s bindings. For
this reason, we {\it delay} the non-determinism until it is needed in
{\it strict contexts} (such as the guard for an {\tt if} a called
procedure, or a numerical primitive application).

@@

\begin{align*}
force(\msto,\clos{\mvar,\mexp,\menv}) &= \{\clos{\mvar,\mexp,\menv}\}\\
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}} \\
\mbox{where } &\mval' \in force(\msto,\mval) \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

The force for the second rule is necessary to keep the same type for
the \(\kfn{}\) continuation, but we would rather use an address. This
introduces (acceptably) more imprecision, but has the same precision as an ANF
analysis\footnote{Exercise for the reader.}.

>>>>

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. Let $wn :
\widehat{State} \times Store \to State$ be the metafunction that adds
a store to a storeless state, $nw$ removes the store, and $c.\msto$
projects the store out of a state $c$. Formally,

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } CS' &= lift(\machstep)(\{ wn(c, \msto) \mid c \in cs\}) \\
              cs' &= \{nw(C') \mid C' \in CS'\} \\
              \msto' &= \bigcup\{C'.\msto \mid C' \in CS'\}
\end{align*}

@@

\begin{align*}
\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev{\mexpi{0},\menv,\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
\\
\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
\ev{\mexp,\menv,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
\\
\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
\\
\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
\\
\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
\ev{\mexp,\menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
\\
\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
\end{align*}

@@

The first problem that we realized with the AAM approach was that much
of the non-determinism is spurious. For example, in a function
application,
\begin{alltt}
(f x y)
\end{alltt}

where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
only to be joined back together in the application position. We would see transitions of the following shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

@@

\begin{align*}
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\}\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}
